---
title: "Lab 3 Hamaniuk-Horetska-Prohorov"
output: html_notebook
---

```{r}
install.packages("tinytex")
tinytex::install_tinytex()
```

\
Lab assignment 3
Parameter estimation and Unbiasedness of Estimator

Work breakdown:

Roman Prohorov is doing Task1 and his efforts are 33.33 %.\
Mariia Hamaniuk is doing Task2 and her efforts are 33.33 %.\
Daryna Horetska is doing Task3 and summary and her efforts are 33.33 %.
*Ctrl+Alt+I*.

------------------------------------------------------------------------

**Task 1**

**Problem formulation and discussion** (what is a reasonable question to discuss):
**R code** with comments:

```{r}

```

**The statistics obtained** (like sample mean or anything else you use to complete the task) as well as histograms
**Justification** of your solution (e.g. refer to the corresponding theorems from probability theory);
**Conclusion** (e.g. how reliable your answer is, if it agrees with common sense expectations).

------------------------------------------------------------------------

**Task 2**

**Problem formulation and discussion** (what is a reasonable question to discuss): \
**R code** with comments:

```{r}
```

**The statistics obtained** (like sample mean or anything else you use to complete the task) as well as histograms\
**Justification** of your solution (e.g. refer to the corresponding theorems from probability theory); **Conclusion** (e.g. how reliable your answer is, if it agrees with common sense expectations).

------------------------------------------------------------------------

**Task 3**

**Problem formulation and discussion** (what is a reasonable question to discuss): \
**R code** with comments:

```{r}
```

**The statistics obtained** (like sample mean or anything else you use to complete the task) as well as histograms \
**Justification** of your solution (e.g. refer to the corresponding theorems from probability theory); **Conclusion** (e.g. how reliable your answer is, if it agrees with common sense expectations).
